Labeling in Strucplots

### 2005-08-30 trunk/vcd/inst/doc/labeling.Rnw
Try better to have \code not hyphenate.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@445 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
viewport name example in strucplot() fixed.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@451 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
mosaic,assoc,and sieve now also accept structable objects. as.table.structable and plot.structable added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@452 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}

%% \code without `-' ligatures
{\catcode`\-=\active%
  \global\def\code{\bgroup%
    \catcode`\-=\active \let-\codedash%
    \Rd@code}}
\def\codedash{-\discretionary{}{}{}}
\def\Rd@code#1{\texttt{#1}\egroup}

@@

\title{Labeling in Strucplots}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\url{David.Meyer@R-project.org}
}
\maketitle
\sloppy

One of the major enhancements in \pkg{vcd} compared to
\code{mosaicplot()} and \code{assocplot()} in
base {\sf R} is the labeling of strucplots which now
has become much more powerful and flexible. Like the panel, the
shading, the spacing, and the legend, labeling is now completely modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \code{strucplot()} which then draws the labels. The labeling is 
well-separated from the actual plotting that occurs in the panel functions;
it actually only relies on the viewport tree produced by them, and the
`dimnames' attribute of the visualized table. The drawing of the
labels happens after the actual plot has been drawn by the panel function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
As an example, consider \code{labeling\_list()}:

\begin{Sinput}
labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
                          sep = " ", offset = unit(c(2, 2), "lines"),
                          varnames = TRUE, cols = 2, ...) 
  function(d, split_vertical, condvars) {
    \dots
  }
\end{Sinput}

\code{labeling\_list()} is a generating function that takes all parameters
as arguments, and returns the actual labeling function which takes
only three arguments: the dimnames structure of the table, the logical
vector with the splitting information, and the number of conditioning
variables (if any). The labeling function has access to the arguments
of the generating function because in {\sf R}, nested functions are
lexically scoped. In the
following, we describe the three basic modules
(\code{labeling\_text()}, \code{labeling\_list()}, and \code{labeling\_cells()}) and
derived functions that build upon them.

@@

\code{labeling\_text()} is the default for all strucplot displays. It
plots labels in the borders similar to the \code{mosaicplot()}
function in base {\sf R}, but is much more flexible: it is not limited
to 4 dimensions, and the positioning and graphical parameters 
of levels and variable names are quite customizable. In addition,
the problem of overlapping labels can be handled in several ways.

@@

\begin{figure}[h]
\begin{center}
<<defaultfig,fig=TRUE,echo=FALSE>>=
<<default>>
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:defaults}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<clippingfig,fig=TRUE,echo=FALSE>>=
<<clipping>>
@
\caption{The effect of clipping.}
\label{fig:clipping}
\end{center}
\end{figure}

@@

\noindent The \code{abbreviate} argument takes a vector of
integers indicating the number of significant characters the levels should be
abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
is performed using {\sf R}'s \code{abbreviate()} function. Another
possibility is to rotate the levels (see Figure \ref{fig:rotating}):

@@

\begin{figure}[h]
\begin{center}
<<rotatefig,fig=TRUE,echo=FALSE>>=
<<rotate>>
@
\caption{Rotating labels.}
\label{fig:rotating}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<repeatfig,fig=TRUE,echo=FALSE>>=
<<repeat>>
@
\caption{Inhibiting the repetition of levels.}
\label{fig:repeat}
\end{center}
\end{figure}

We now proceed with a few more `cosmetic' features.
A first simple, but effectful modification is to position
all labels and variables left-aligned, which can
also be achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):

@@

\begin{figure}[h]
\begin{center}
<<marginsfig,fig=TRUE,echo=FALSE>>=
<<margins>>
@
\caption{Changes in the margins.}
\label{fig:margins}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
<<boxesfig,fig=TRUE,echo=FALSE>>=
<<boxes>>
@
\caption{Boxes and Clipping.}
\label{fig:boxes}
\end{center}
\end{figure}

\noindent The values to \code{boxes} and \code{clip} are recycled for
all dimensions. The result is pretty close to what the \code{labels\_cboxed()}
wrapper does. Another variant is to put the variable names into the
same line as the levels (see Figure \ref{fig:labbl}):

@@

\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
labels'') is a vector of logicals indicating the side for the
variable names. The resulting layout is basically what \code{labeling\_lboxed()} produces,
and a similar design is used by the \code{doubledecker()}.

@@

\noindent For some data, it might be convenient to combine cell
labeling with border labeling as done by \code{labels\_conditional()}
(see Figure \ref{fig:conditional}):

@@

\section{A simple list of labels:\texttt{labels\_list()}}

>>>>

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}

%% \code without `-' ligatures
\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
\def\restorehyphenation{\hyphenchar\font=`-}
{\catcode`\-=\active%
  \global\def\code{\bgroup%
    \catcode`\-=\active \let-\codedash%
    \Rd@code}}
\def\codedash{-\discretionary{}{}{}}
\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}

\newcommand{\codefun}[1]{\code{#1()}}

@@

\title{Labeling in Strucplots}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\email{David.Meyer@R-project.org}
}
\maketitle
\sloppy

One of the major enhancements in \pkg{vcd} compared to
\codefun{mosaicplot} and \codefun{assocplot} in
base {\sf R} is the labeling in the strucplot framework which now
has become much more powerful and flexible. Like the panel, the
shading, the spacing, and the legend, labeling is now completely modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \codefun{strucplot} which then draws the labels (generating
functions are discussed in a separate vignette). The labeling is 
well-separated from the actual plotting that occurs in the panel functions;
it actually only relies on the viewport tree produced by them, and the
`dimnames' attribute of the visualized table. The drawing of the
labels happens after the actual plot has been drawn by the panel function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
% As an example, consider \codefun{labeling\_list}:

% \begin{Sinput}
% labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
%                           sep = " ", offset = unit(c(2, 2), "lines"),
%                           varnames = TRUE, cols = 2, ...) 
%   function(d, split_vertical, condvars) {
%     \dots
%   }
% \end{Sinput}

% \codefun{labeling\_list} is a generating function that takes all parameters
% as arguments, and returns the actual labeling function which takes
% only three arguments: the dimnames structure of the table, the logical
% vector with the splitting information, and the number of conditioning
% variables (if any). The labeling function has access to the arguments
% of the generating function because in {\sf R}, nested functions are
% lexically scoped. 
In the following, we describe the three basic modules
(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
derived functions that build upon them.

@@

\codefun{labeling\_text} is the default for all strucplot displays. It
plots labels in the borders similar to the \codefun{mosaicplot}
function in base {\sf R}, but is much more flexible: it is not limited
to 4 dimensions, and the positioning and graphical parameters 
of levels and variable names are quite customizable. In addition,
the problem of overlapping labels can be handled in several ways.

@@

\begin{figure}[p]
\begin{center}
<<defaultfig,fig=TRUE,echo=FALSE>>=
<<default>>
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:defaults}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<clippingfig,fig=TRUE,echo=FALSE>>=
<<clipping>>
@
\caption{The effect of clipping.}
\label{fig:clipping}
\end{center}
\end{figure}

@@

\noindent The \code{abbreviate} argument takes a vector of
integers indicating the number of significant characters the levels should be
abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
is performed using the \codefun{abbreviate} function in base {\sf R}. Another
possibility is to rotate the levels (see Figure \ref{fig:rotating}):

@@

\begin{figure}[p]
\begin{center}
<<rotatefig,fig=TRUE,echo=FALSE>>=
<<rotate>>
@
\caption{Rotating labels.}
\label{fig:rotating}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<repeatfig,fig=TRUE,echo=FALSE>>=
<<repeat>>
@
\caption{Inhibiting the repetition of levels.}
\label{fig:repeat}
\end{center}
\end{figure}

We now proceed with a few more `cosmetic' features.
A first simple, but effectful modification is to position
all labels and variables left-aligned, which can
also be achieved by using \codefun{labeling\_left} (see Figure \ref{fig:left}):

@@

\begin{figure}[p]
\begin{center}
<<marginsfig,fig=TRUE,echo=FALSE>>=
<<margins>>
@
\caption{Changes in the margins.}
\label{fig:margins}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<boxesfig,fig=TRUE,echo=FALSE>>=
<<boxes>>
@
\caption{Boxes and Clipping.}
\label{fig:boxes}
\end{center}
\end{figure}

\noindent The values to \code{boxes} and \code{clip} are recycled for
all dimensions. The result is pretty close to what the \codefun{labels\_cboxed}
wrapper does. Another variant is to put the variable names into the
same line as the levels (see Figure \ref{fig:labbl}):

@@

\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
labels'') is a vector of logicals indicating the side for the
variable names. The resulting layout is basically what \codefun{labeling\_lboxed} produces,
and a similar design is used by the \codefun{doubledecker}.

@@

\noindent For some data, it might be convenient to combine cell
labeling with border labeling as done by \codefun{labels\_conditional}
(see Figure \ref{fig:conditional}):

@@

\section{A simple list of labels: \texttt{labels\_list()}}
